---
title: "Week 3/4 Homework"
author: "Arin Vansomphone"
date: "`r Sys.Date()`"
output: html_document
---
### 1. Importing the datasets:
```{r message=FALSE, warning=FALSE}
# loading the packages
library(tidyverse)
library(here)
library(gapminder)

# importing the datasets
demoinfo <- read_csv("accord_key.csv")
bloodpress <- read_csv("bloodpressure.csv")
bloodsugar <- read_csv("hba1c.csv")
```

### 2. Describing the datasets
1. According to the label sheet, MaskID is the number given to each participant in the database. Looking at the `demoinfo` database, it says there are 10,251 entries, so there must be 10,251 participants in each of the three databases. Let's find the gender ratio now:

```{r message=TRUE, warning=FALSE}
demoinfo %>%
  group_by(female) %>%
  summarize(number = n(),
            percent = 100 *number / 10251)
```

The database population is roughly 61.4% male and 38.6% female.

Grouping by ethnicity now:

```{r}
demoinfo %>%
  group_by(raceclass) %>%
  summarize(number = n(),
            percentage = 100 * number / 10251)
```

The database consists of around 19.1% Black people, 7.2% Hispanic people, 11.4% Other people, and 62.4% White people.

2. To find the mean and interquartile numbers for the blood pressure and sugar of each individual, we need to use the summarize function again.

```{r}
# blood pressure numbers 
bloodpress %>%
  group_by(MaskID) %>%
  summarize(mean = mean(sbp),
            IQR = IQR(sbp))

# blood sugar numbers
bloodsugar %>%
  group_by(MaskID) %>%
  summarize(mean = mean(hba1c),
            IQR = IQR(hba1c))
```

### 3. Extracting baseline SBP, DBP, and hba1c numbers
To extract summary stats on the baseline SBP, DBP, and hba1c numbers, we input `filter(Visit == "BLR")` into our piped function.

Baseline SBP:
```{r}
bloodpress %>%
  filter(Visit == "BLR") %>%
  summarise(mean = mean(sbp),
            std = sd(sbp),
            IQR = IQR(sbp),
            median = median(sbp))
```

Baseline DBP:
```{r}
bloodpress %>%
  filter(Visit == "BLR") %>%
  summarise(mean = mean(dbp),
            std = sd(dbp),
            IQR = IQR(dbp),
            median = median(dbp))
```

Baseline hba1c:
```{r}
bloodsugar %>%
  filter(Visit == "BLR") %>%
  summarise(mean = mean(hba1c),
            std = sd(hba1c),
            IQR = IQR(hba1c),
            median = median(hba1c))
```

Now that we've found the numbers for the entire population, let's find the summary stats for select groups of the population. To do this, we need to merge the `bloodpress` and `bloodsugar` dataframes into the `demoinfo` dataframe so we can have the ethnicity and gender information.
```{r}
bigdataset <- demoinfo %>%
  inner_join(bloodpress, by = "MaskID")

bigdataset <- bigdataset %>%
  inner_join(bloodsugar, by = "MaskID")
```

Let's first find the summary stats based on gender:
SBP:
```{r}
bigdataset %>%
  filter(Visit.x == "BLR") %>%
  group_by(female) %>%
  summarize(mean = mean(sbp),
            std = sd(sbp),
            median = median(sbp),
            IQR = IQR(sbp))
```

DBP:
```{r}
bigdataset %>%
  filter(Visit.x == "BLR") %>%
  group_by(female) %>%
  summarize(mean = mean(dbp),
            std = sd(dbp),
            median = median(dbp),
            IQR = IQR(dbp))
```

hba1c:
```{r}
bigdataset %>%
  filter(Visit.y == "BLR") %>%
  group_by(female) %>%
  summarize(mean = mean(hba1c),
            std = sd(hba1c),
            median = median(hba1c),
            IQR = IQR(hba1c))
```

Now based on ethnicity:
SBP:
```{r}
bigdataset %>%
  filter(Visit.x == "BLR") %>%
  group_by(raceclass) %>%
  summarize(mean = mean(sbp),
            std = sd(sbp),
            median = median(sbp),
            IQR = IQR(sbp))
```

DBP:
```{r}
bigdataset %>%
  filter(Visit.x == "BLR") %>%
  group_by(raceclass) %>%
  summarize(mean = mean(dbp),
            std = sd(dbp),
            median = median(dbp),
            IQR = IQR(dbp))
```

hba1c:
```{r}
bigdataset %>%
  filter(Visit.y == "BLR") %>%
  group_by(raceclass) %>%
  summarize(mean = mean(hba1c),
            std = sd(hba1c),
            median = median(hba1c),
            IQR = IQR(hba1c))
```

### 4. Understanding the trend of SBP, DBP, and hba1c over visits
1. 
```{r}
bloodpress %>%
  group_by(Visit) %>%
  summarize(meanSBP = mean(sbp))
```

2. 
```{r}
bloodpress %>%
  group_by(Visit) %>%
  summarize(meanSBP = mean(sbp)) %>%
  ggplot(aes(x = Visit, y = meanSBP)) + 
  geom_point() + 
  coord_flip() +
  labs(title = "Mean SBP per Visit",
       x = "Mean SBP",
       y = "Visit")
```

4. Repeating the procedure for DBP and hba1c:
DBP:
```{r}
bloodpress %>%
  na.omit() %>%
  group_by(Visit) %>%
  summarize(meanDBP = mean(dbp)) %>%
  ggplot(aes(x = Visit, y = meanDBP)) + 
  geom_point() + 
  coord_flip() +
  labs(title = "Mean DBP per Visit",
       x = "Mean DBP",
       y = "Visit")
```

hba1c: 
```{r}
bloodsugar %>%
  na.omit() %>%
  group_by(Visit) %>%
  summarize(meanhba1c = mean(hba1c)) %>%
  ggplot(aes(x = Visit, y = meanhba1c)) + 
  geom_point() + 
  coord_flip() +
  labs(title = "Mean hba1c per Visit",
       x = "Mean hba1c",
       y = "Visit")
```

### 5. Bonus
For the bonus we need to create a new variable named "glycemic_arm" that sorts the arm types into 2 categories. To accomplish this we will create a set of values corresponding to each type of glycemic arm and ask R to assign values based on the set.
```{r warning=FALSE}
# converting arm to character type and creating values for 0
bigdataset$arm <- as.character(bigdataset$arm)
glycemicarm0 = c("1", "2", "5", "6")

# creating the new variable and assigning values
bigdataset$glycemic_arm = "1"
bigdataset$glycemic_arm[bigdataset$arm %in% glycemicarm0] = "0"
```

Plotting the mean profile plot grouped by glycemic arm is nearly the same as we did previously:
```{r warning=FALSE}
bigdataset %>%
  group_by(Visit.y, glycemic_arm) %>%
  summarize(mean = mean(hba1c)) %>%
  ggplot(aes(x = Visit.y, y = mean, color = glycemic_arm)) +
  geom_point() + 
  coord_flip() +
  labs(title = "Mean Profile Plot of hba1c Grouped by Glycemic Arm",
       x= "Visit",
       y = "Mean")
```

