---
title: "Week 3 Notes"
author: "Arin Vansomphone"
date: "`r Sys.Date()`"
output: html_document
---
# Wrangle Notes
## Tibbles
```{r}
library(tidyverse)
```
Tibbles are like dataframes, but upgraded with more features

### Exercises
1. 
```{r}
mtcars
```
You can tell if an object is a tibble by looking at the description.

2. 
```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

df2 <- tibble(abc = 1, xyz = "a")
df2$x
df2[, "xyz"]
df2[, c("abc", "xyz")]
```
There isn't much difference between the tibble and dataframe, except for the description. However, the matching operations are much more strict and orderly on tibbles, as it shows you the multiple steps of each operation on the object. The dataframe only outputs one dataframe.

3. `tibblename$mpg`

4.
```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

# 4.1
annoying$`1`

# 4.2 
annoying %>%
  ggplot(aes(x = `1`, y = `2`)) +
  geom_point()

# 4.3 
annoying <- annoying %>%
  mutate(`3` = `2` / `1`)

# 4.4
colnames(annoying) <- c("one", "two" , "three")
```

5. 
```{r}
tibble::enframe(annoying$two)
```
The operation gives the values of the column and the order that they appear in. It could be useful when ordering least to greatest values or vice versa.

6. The `n_extra` argument determines how many more columns to print for.

## Data import
* Always use `read_csv()`.
* The `parse_*()` function can take a character vector and return a more specialized vector like a logical, integer, or date (basically changing the type). Also make sure the argument is a vector!
* `parse_number()` is useful when you want to ignore non-numeric characters before and after the number, such as currencies and percentages.
* `parse_date()` expects a four digit year, the month, then the day.
* `write_csv()` writes the data back into disk.

## Tidy data
* Rules for making a dataset tidy:
1. Each variable must have its own column
2. Each observation must have its own row
3. Each value must have its own cell
* Creating tidy data allows vectorized functions like mutate and summarize to perform extremely well.
* Use `pivot_longer()` when there are variable values as names of the columns. Generally, it increases the rows and decreases the columns.
* Use `pivot_wider()` when an observation is scattered across multiple rows. Decreases the rows and increases the columns.
* `separate()` pulls apart a column and turns it into multiple. Used for when there are multiple values in one cell.
* `unite()` is the exact opposite of `separate()`.
* Explicitly missing data: data flagged with NA
* Implicitly missing data: data just missing from the dataset
* You can make implicitly missing data explicit by using the `complete()` command, which will find all the unique combinations and fill in NA where necessary.

## Relational data
Relational data: multiple tables of data in which the relations between them are important

Three families of verbs to work with relational data:
* Mutating joins, adding new variables to a data frame from matching observations in another
* Filtering joins, filtering observations from one data frame based on whether or not they match an observation in the other table
* Set operations, treating observations as if they were set elements

* Remember that each relation always concerns a pair of tables.
* A key is a variable that uniquely identifies an observation (Mask ID for example)

## Strings
Strings are written with " or ', but the double quote is always preferred.
```{r}
library(htmlwidgets)
x <- c("apple", "banana", "pear")
str_view(x, "an")
```
* Use regular expressions to find a specific phrase/character in the dataset

## Factors 
The two factor operations:
1. Modifying the factor order. Use `fct_reorder` if you want a rearrangement from least to greatest.
2. Modifying factor levels. Use `fct_recode` to change the factor level, group values together, or separate them out.

## Dates and times
Three types of date data:
1. A date
2. A time within a day
3. A date-time, which is a combination of the two

How to convert to date data from strings (need the package lubridate):
```{r}
library(lubridate)
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")

# including time as well
ymd_hms("2017-01-31 20:11:59")
```
Getting components of the date data:
```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
mday(datetime)

yday(datetime)
wday(datetime)
```
* To round the date to a specific unit of time, use `floor_date()`, `round_date()`, or `ceiling_date()`.
* In R, when you subtract two dates, you get a difftime object. To avoid some pain, use the `as.duration` function from the lubridate package.

#Program notes
## Pipes
* The point of a pipe is to help you read and understand your code easier
* Pipes work by performing a lexical transformation

Unfortunately, pipes won't work for 2 classes of functions:
1. Functions that use the current environment
2. Functions that use lazy evaluation (A system in which the function is evaluated only when needed)

When not to use the pipe;
* The pipe is longer than around ten steps
* When there are multiple objects being manipulated and multiple inputs/outputs
* When trying to make a directed graph with a complex dependency structure

## Functions
* Consider writing a function when you've copied and pasted a block of code more than twice

Three steps to creating a function:
1. Pick a good name for the function
2. List the inputs or arguments to the function
3. Place the code for the function in a `{` block
Example:
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))
```
* If you have a family of functions that do similar things, make the prefix similar rather than the suffix:
```{r}
# Good
input_select()
input_checkbox()
input_text()

# Not so good
select_input()
checkbox_input()
text_input()
```
* if statements:
```{r}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```
* elseif statements:
```{r}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  # 
}
```

* There are two types of function arguments: data and detail. Data is the actual input of the function, while detail controls the details of the computation
* Function arguments in R are lazily evaluated, meaning that they will only be evaluated if they are called
* R will look in the environment of where the function is defined

## Vectors
Two types of vectors:
* Atomic vectors, which have six subtypes: logical, integer, double, character, complex, and raw
* Lists

* Atomic vectors are homgeneous, while lists can be heterogeneous
* Every vector has two key properties: `typeof()` and `length()`
* Doubles are approximations, like floats in Julia
* Change the type of the vector with `as.logical()`, `as.integer()`, `as.double()`, `as.character()`
* Use `[` to subset

* Lists can contain other lists

* Attribute: named list of vectors that can be attached to any object
```{r}
x <- 1:10
attr(x, "greeting")
#> NULL
attr(x, "greeting") <- "Hi!"
attr(x, "farewell") <- "Bye!"
attributes(x)
#> $greeting
#> [1] "Hi!"
#> 
#> $farewell
#> [1] "Bye!"
```
* Three important types of attributes:
  1. Names
  2. Dimensions
  3. Class

Augmented vectors are vectors with an additional attribute, class. There are four main types of augmented vectors:
  * Factors
  * Dates
  * Date-times 
  * Tibbles

__Important to remember:__ R is a vectorized language, so almost all operations will use vectors! Think about
that when coding!

## Iteration
Iteration helps you do the same thing to multiple inputs, columns, or datasets

* `seq_along()` is a better verson of `1:length()`. Use it like this:
```{r}
output <- vector("double", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
```


Every for loop has three essential parts:
  1. Output 
  2. Sequence (how long to run the loop for)
  3. Body (the actual code)

For individual elements in a database, use [[i]] (this is the mistake you made in the glycemic arm problem)

Cool function for getting summary stats on a column:
```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
col_summary(df, median)
#> [1] -0.51850298  0.02779864  0.17295591 -0.61163819
col_summary(df, mean)
#> [1] -0.3260369  0.1356639  0.4291403 -0.2498034
```


* `map` functions could also be useful for doing operations on all the elements in a vector