---
title: "Week 8 Work"
author: "Arin Vansomphone"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

# 1,5 AG Analysis
Loading the packages:
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
```

Loading the data:
```{r}
bloodpressure <- read_csv("bloodpressure.csv")
hba1c <- read_csv("hba1c.csv")
AG <- read_xlsx("UCLA Samples-2022-07-19 Summary.xlsx", sheet = "Summary",
                skip = 7) %>%
  rename(BSI_ID = `Sample ID`)
linkfile <- read_csv("zhou_7816_linkfile.csv")
accord_key <- read_csv("accord_key.csv")
microvascularoutcomes <- read_csv("microvascularoutcomes.csv")
cvdoutcomes <- read_csv("cvdoutcomes.csv")
f07_baseline <- read_csv("f07_baselinehistoryphysicalexam.csv")
otherlabs <- read_csv("otherlabs.csv")
bloodpressure <- read_csv("bloodpressure.csv")
lipids <- read_csv("lipids.csv")
AG
```

Tidying the data:
```{r}
AG <- AG %>% #linking MASKID to Batch ID
  full_join(linkfile, by = "BSI_ID") %>% 
  filter(!is.na(MASKID) ) %>% 
  rename(MaskID = MASKID,
         Box = `R2020 000036`)
AG
```
```{r}
# creating a BatchID column
BatchIDs <- c("R2020 000036", "R2020 000037", "R2020 000038", "R2020 000039", 
             "R2020 000040", "R2020 000041", "R2020 000042", "R2020 000043")
index = c(1, which(AG$Box %in% BatchIDs), 4001)
BatchID_col = rep(BatchIDs, diff(index))
AG <- AG %>% mutate(BatchID = BatchID_col) %>%
  relocate(BatchID, .before = Box) %>%
  relocate(MaskID, .after = BSI_ID)
AG
```
```{r}
# replacing errors in Box
AG["Box"][AG["Box"] == "R2020 000037"] <- "Box 1"
AG["Box"][AG["Box"] == "R2020 000038"] <- "Box 1"
AG["Box"][AG["Box"] == "R2020 000039"] <- "Box 1"
AG["Box"][AG["Box"] == "R2020 000040"] <- "Box 1"
AG["Box"][AG["Box"] == "R2020 000041"] <- "Box 1"
AG["Box"][AG["Box"] == "R2020 000042"] <- "Box 1"
AG["Box"][AG["Box"] == "R2020 000043"] <- "Box 1"
AG
```
```{r}
# joining hba1c with the AG dataframe
AG <- AG %>%
  inner_join(hba1c, by = "MaskID")
AG
```
```{r}
# joining other dataframes
AG <- AG %>%
 left_join(accord_key, by= "MaskID") %>%
 left_join(f07_baseline, by = c("MaskID", "Visit")) %>%
 left_join(lipids, by = c("MaskID", "Visit")) %>%
 left_join(bloodpressure, by = c("MaskID", "Visit")) %>%
 left_join(otherlabs, by = c("MaskID", "Visit"))
AG
```
```{r}
# creating glycemic, bp, and lipid arm variables
AG <- AG %>%
mutate(glycemic_arm = ifelse(arm==1|arm==2|arm==5|arm==6,0,1),
         bp_arm = ifelse(arm==3 | arm ==1, 1, 0),
         lipid_arm = ifelse(arm==7 | arm == 5, 1, 0))
AG
```
```{r}
# creating the kidney variable
baseline_ualb <- otherlabs %>% 
  filter(Visit == "BLR") %>% 
  mutate(kidney = ifelse(gfr < 45 | ualb > 300, 1, 0)) %>% 
  select(MaskID, -Visit, kidney)
baseline_ualb
```
```{r}
# joining the kidney variable with the main dataframe
AG <- AG %>%
  inner_join(baseline_ualb, by = "MaskID")
AG
```
```{r}
# mutating other biomarkers
AG <- AG %>%
  mutate(bmi = wt_kg / ((ht_cm / 100) ^ 2),
         )
```
```{r}
# selecting and filtering the data that we need
AG <- AG %>%
  select(BatchID, Box, MaskID, Visit, Mean, hba1c,
         baseline_age, glycemic_arm, bp_arm, lipid_arm,
         female, raceclass, cvd_hx_baseline, yrsdiab, 
         hartfail, eyedisea, smokelif, kidney,
         fpg, gfr, bmi, uacr,
         sbp, dbp, hdl, ldl,
         trig, chol) %>%
  rename(gender = female)
AG
```

### Are there any differences between hba1c levels at 12 months across batches?
Plotting the hba1c levels at 12 months (grouped by `BatchID`):
```{r}
AG %>%
  filter(Visit == "F12") %>%
  group_by(BatchID) %>%
  summarize(Mean = mean(hba1c, na.rm = TRUE),
            Median = median(hba1c, na.rm = TRUE),
            SD = sd(hba1c, na.rm = TRUE)) %>%
  ggplot(aes(x = BatchID, y = Mean)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_fixed(ylim = c(5, 10))
  labs(title = "Mean hba1c levels at 12 months (grouped by Batch ID)",
       x = "Batch IDs",
       y = "hba1c")
```
```{r}
AG %>%
  group_by(BatchID) %>%
  summarize(Mean = mean(hba1c, na.rm = TRUE),
            Median = median(hba1c, na.rm = TRUE),
            SD = sd(hba1c, na.rm = TRUE))
```

According to the data, there are no differences in hba1c levels at 12 months across batches. 

### What are the factors at baseline that contribute to 1,5 AG?
__Computing the pearson correlation coefficients for continuous covariates:__
```{r}
library(corrplot)
library(RColorBrewer)
library(xtable)
library(DT)
```
```{r}
AG <- AG %>% mutate(flag6 = factor(ifelse(Mean >= 6, 1, 0), 
                                   labels = c("1,5AG < 6", "1,5AG ≥ 6")),
                    flag6 = relevel(flag6, ref = "1,5AG ≥ 6"))

correlationvars <- AG %>%
  rename(`1,5 AG` = Mean,
         HbA1c = hba1c,
         Age = baseline_age,
         Yrsdiab = yrsdiab,
         Glucose = fpg,
         BMI = bmi,
         SBP = sbp,
         DBP = dbp,
         HDL = hdl,
         LDL = ldl,
         Trig = trig,
         Chol = chol,
         uACR = uacr,
         eGFR = gfr) %>%
  select(flag6, `1,5 AG`, HbA1c, Age, Yrsdiab, Glucose, BMI, SBP, DBP, HDL, LDL, 
         Trig, Chol, uACR, eGFR)
correlationvars
```
```{r}
corrplot(cor(correlationvars[-1], use = "complete.obs"), 
         type = "upper", order = "original", diag = FALSE,
         col=brewer.pal(n = 10, name = "RdYlBu"), 
         title = "Correlation Plot Between Baseline Covariates", mar=c(0,0,1,0))
```

Creating a table for the values:
```{r}
corvarstable <- correlationvars %>%
  na.omit() %>%
  select(-flag6)

mcor <- round(cor(corvarstable),2)

lower<- mcor
lower[lower.tri(mcor, diag=TRUE)]<-""
lower <- as.data.frame(lower)

DT::datatable(lower, 
         extensions = c('FixedColumns',"FixedHeader"),
          options = list(scrollX = TRUE, 
                         paging=FALSE,
                         fixedHeader=TRUE))
```

None of these factors show a particularly strong correlation to 1,5 AG.

__Univariate regression for categorical variables:__

Merging data:
```{r}
tindpt <- AG %>%
  select(MaskID, Mean, BatchID,
         baseline_age,
         glycemic_arm, bp_arm, lipid_arm, gender, raceclass, cvd_hx_baseline,
         yrsdiab, hartfail, eyedisea, smokelif, kidney) %>% 
  mutate(glycemic_arm = factor(glycemic_arm, levels = c(1,0), labels = c("Yes", "No")),
         bp_arm = factor(bp_arm, levels = c(1,0), labels = c("Yes", "No")),
         lipid_arm = factor(lipid_arm, levels = c(1,0), labels = c("Yes", "No")),
         gender = factor(gender, levels = c(1,0), labels = c("Yes", "No")),
         raceclass = factor(raceclass, levels  = c("Black", "Hispanic", "White", "Other")),
         cvd_hx_baseline = factor(cvd_hx_baseline, levels = c(1,0), labels = c("Yes", "No")),
         hartfail = factor(hartfail, levels = c(1,2), labels = c("Yes", "No")),
         eyedisea = factor(eyedisea, levels = c(1,2), labels = c("Yes", "No")),
         smokelif = factor(smokelif, levels = c(1,2), labels = c("Yes", "No")),
         kidney = factor(kidney, levels = c(1,0), labels = c("Yes", "No"))) %>%
  mutate(gender = relevel(gender, ref = "No"),
         cvd_hx_baseline = relevel(cvd_hx_baseline, ref = "No"),
         glycemic_arm = relevel(glycemic_arm, ref = "No"),
         bp_arm = relevel(bp_arm, ref = "No"),
         lipid_arm = relevel(lipid_arm, ref = "No"),
         eyedisea = relevel(eyedisea, ref = "No"),
         kidney = relevel(kidney, ref="No"),
         hartfail = relevel(hartfail, ref = "No"),
         smokelif = relevel(smokelif, ref = "No"))
```

Fitting a linear regression:
```{r}
fit <- lm(log(Mean) ~ . - MaskID, tindpt)
summary(fit)
```

### What baseline medications contribute to 1,5 AG?
Loading the data:
```{r}
concomitantmeds <- read_csv("concomitantmeds.csv")
```

Filtering to baseline and creating the `insulin` variable:
```{r}
concomitantmeds_baseline <- concomitantmeds %>% 
  filter(Visit == "BLR") %>%
  mutate(insulin = ifelse(nphl_insulin==1|reg_insulin==1|la_insulin==1 | 
                         othbol_insulin==1 |premix_insulin == 1, 1, 0))
```

Selecting the desired variables:
```{r}
medications <- concomitantmeds_baseline %>%
  inner_join(AG, by = c("MaskID", "Visit")) %>% # for 1,5 AG 
  select(MaskID, Mean, a2rb, acei, fibrate, statin, sulfonylurea, biguanide, 
         meglitinide, tzd, insulin)
```

Merging data:
```{r}
tindpt2 <- medications %>%
  mutate(a2rb = factor(a2rb, levels = c(1,0), labels = c("Yes", "No")),
         acei = factor(acei, levels = c(1,0), labels = c("Yes", "No")),
         fibrate = factor(fibrate, levels = c(1,0), labels = c("Yes", "No")),
         statin = factor(statin, levels = c(1,0), labels = c("Yes", "No")),
         sulfonylurea = factor(sulfonylurea, levels = c(1,0), labels = c("Yes", "No")),
         biguanide = factor(biguanide, levels = c(1,0), labels = c("Yes", "No")),
         meglitinide = factor(meglitinide, levels = c(1,0), labels = c("Yes", "No")),
         tzd = factor(tzd, levels = c(1,0), labels = c("Yes", "No")),
         insulin = factor(insulin, levels = c(1,0), labels = c("Yes", "No"))) %>%
  mutate(a2rb = relevel(a2rb, ref = "No"),
         acei = relevel(acei, ref = "No"),
         fibrate = relevel(fibrate, ref = "No"),
         statin = relevel(statin, ref = "No"),
         sulfonylurea = relevel(sulfonylurea, ref = "No"),
         biguanide = relevel(biguanide, ref = "No"),
         meglitinide = relevel(meglitinide, ref="No"),
         tzd = relevel(tzd, ref = "No"),
         insulin = relevel(insulin, ref = "No"))
```

Fitting a linear regression:
```{r}
fit2 <- lm(log(Mean) ~ . - MaskID, tindpt2)
summary(fit2)
```